Ensembler: Enabling high-throughput molecular simulations at the superfamily scale

### 2015-02-06 manuscript/ms.tex
Wrote introduction

% The figures are in a figures/ subdirectory.
\graphicspath{{figures/}}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% DESIGN AND IMPLEMENTATION
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Design and Implementation}
\label{section:design}

>>>>

% The figures are in a figures/ subdirectory.
% \graphicspath{{figures/}}

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% DESIGN AND IMPLEMENTATION
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Design and Implementation}

% Python
% Can be used via command-line tool Ensembler, or via the Python API

Ensembler is written in Python, and can be used via a command-line tool ({\tt
ensembler}), or via the flexible Python API.

% Pipeline architecture, with a series of stages which must be performed in order.

The Ensembler modeling pipeline entails a series of stages which are performed
in a defined order. A visual overview of the pipeline is shown in Fig.
\ref{figpipeline}, and a detailed description follows.

\begin{figure*}[h]
  \centering
    \includegraphics[width=0.8\textwidth]{pipeline}
  \caption{Ensembler pipeline}
  \label{figpipeline}
\end{figure*}

\subsubsection{Target selection}

The output from this stage is simply a FASTA-formatted sequence file containing
the selected target sequences and identifiers. The {\tt ensembler} command-line
tool provides methods for selecting targets from either UniProt (a freely
accessible resource for protein sequence and functional
data---\href{http://www.uniprot.org/}{uniprot.org}) or TargetExplorer (a
database framework for aggregating various types of biological data; work to be
published). This allows the user to easily select a single protein, many
proteins, or an entire superfamily. Alternatively, the targets file can be
generated using any other software, and stored at the appropriate filepath.

The method for selecting targets from UniProt is described here. A query string
is required as input, using the same syntax as the search function on the
UniProt website. For example, {\tt'domain:"Protein kinase" AND taxonomy:9606
AND reviewed:yes'} would select all human protein kinases which have been
reviewed by a human curator. Ensembler is designed to work with protein
\emph{domains}, rather than full-length proteins, and the desired protein
domain(s) can be selected using a regular expression. For example, the string
{\tt'\^{}Protein kinase(?!; truncated)(?!; inactive)'} would match domains
annotated as "Protein kinase", "Protein kinase; 1" or "Protein kinase; 2", but
would exclude the domains "Protein kinase; truncated" and "Protein kinase;
inactive". The program then extracts sequences and identifiers from the UniProt
data, and saves these to a FASTA-format text file.

% The selected targets are
% given identifiers of the form {\tt [UniProt mnemonic]\_D[domain index]}, which
% consists of the UniProt name for the target protein and a 0-based index for the
% domain (since a single target protein may contain multiple domains of
% interest). Example: {\tt EGFR\_HUMAN\_D0}.

\subsubsection{Template selection}

As for target selection, the {\tt ensembler} tool provides methods for
selecting templates from various resources---UniProt, the Protein Data Bank
(PDB; \href{http://www.pdb.org}{pdb.org}) or a TargetExplorer database.  From
the user perspective, selection from UniProt proceeds in a similar fashion as
described above. The returned data for each UniProt entry includes an
up-to-date list of PDB structures and their residue spans. PDB files are
downloaded from the PDB for structures which include the desired domain. Data
from the SIFTS service
(\href{http://www.ebi.ac.uk/pdbe/docs/sifts/}{www.ebi.ac.uk/pdbe/docs/sifts})
(CITE: Velankar Nucleic Acids Res 2013), which provides residue-level mappings
between PDB and UniProt entries, is then used to filter out PDB chains with $<$
70\% resolved residues within the domain span. Template sequences and
structures are then extracted and written as FASTA-format sequence and
PDB-format coordinate files respectively.

Selection from the PDB simply requires specifying a list of PDB IDs. These are
matched to UniProt entries via the SIFTS service, and the same procedure is
then followed to extract template sequences and structures.

Unresolved template loops can optionally be remodeled with a kinematic closure
algorithm, which is provided via the loopmodel tool of the Rosetta software
suite (CITE: Rosetta and/or loopmodel). This tends to provide higher-quality
models following the subsequent modeling process.

\subsubsection{Modeling}

In this stage, models are generated for each target-template pair, using the
Modeller automodel function (CITE: Modeller), which implements comparative
structure modeling by satisfaction of spatial restraints (CITE: Sali Blundell J
Mol Biol 1993; Fiser Sali Prot Sci 9 2000). Modeller requires the user to first
provide a target-template sequence alignment. This is implemented in Ensembler
using the BioPython {\tt pairwise2} module (CITE: BioPython)---which uses a
dynamic programming algorithm---with the PAM 250 scoring matrix of Gonnet
\textit{et al} (CITE: Gaston Gonnet Science 1992).

Non-unique models are then filtered out using structural similarity-based
clustering. The mdtraj (CITE: mdtraj) Python library is used to calculate RMSD
with a fast quaternion characteristic polynomial (QCP) implementation, and the
leader algorithm is then used to populate clusters. A minimum distance cutoff
(default: 0.6 \AA) is used to retain only a single model per cluster.

\subsubsection{Refinement}

Models are then refined with a steepest descent energy minimization and a short
molecular dynamics (MD) simulation with implicit solvent. This is implemented
using the OpenMM molecular simulation toolkit (link and CITE: OpenMM), chosen
for its flexible Python API, and high performance GPU-acclerated simulation
code. The Amber99SB-ILDN force field is used (CITE: amber99sbildn refs) with a
modified generalized Born solvent model (GBSA-OBC) (CITE: GBSA-OBC). The
simulation is run for a default of 100 ps. This refinement process helps to
prepare models for subsequent production simulation, and also helps to filter
out poor quality models.

Ensembler also provides optional routines for solvating models with explicit
solvent and performing a second MD refinement. The solvation step solvates each
model for a given target with the same number of waters, as this is (currently)
a requirement for building MSMs from multiple independent MD trajectories. The
target number of waters is selected by first solvating each model with a
specified padding distance (default: 10 \AA), then taking a percentile value
from the distribution (default: 68th percentile). Models are resolvated with
the target number of waters by first solvating with zero padding, then
incrementally increasing the box size and resolvating until the target is
exceeded, then finally deleting sufficient waters to match the target value.
The explicit solvent MD simulation is also implemented using OpenMM, with the
Amber99SB-ILDN force field and TIP3P water.

\subsubsection{Packaging}

Finally, Ensembler provides a packaging module, which can be used to compress
models in preparation for data transfer, or to prepare models with the
appropriate directory and file structure for subsequent production simulations
on the distributed computing platform Folding@Home (CITE: F@H).

\subsubsection{Other features}

% quickmodel function performs the entire pipeline in one go, for a single target and a small number of templates.
The command-line tool also provides a {\tt quickmodel} function, which performs
the entire Ensembler pipeline for a single target with a small number of
templates. For larger numbers of models (such as entire protein families), the
main pipeline functions should be used. The modeling and refinement functions
use MPI to trivially parallelize computation across each model (or across each
template, in the case of the loop reconstruction code).

Each pipeline function also outputs a metadata file, which helps to link data
to the software version used to generate it (both Ensembler and its
dependencies), and also provides timing and performance information, and other
data such as hostname.

% ? The API is highly flexible, allowing control over many important parameters. Appropriate choices have been set as default values.

% TODO where do these go?
% Distribution: conda package or github source
% Software is open source and designed to be extensible as far as possible. For example, it would be very simple to implement further options for providing target and template structures, or for packaging models.

\label{section:design}
